---
title: "Getting started with blink"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with blink}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# The examples build the odin2 model and use malariasimulation + postie (both in
# Suggests). They are runnable as written; to keep package checks fast and
# dependency-light, chunks only evaluate when those packages are installed.
can_run <- requireNamespace("malariasimulation", quietly = TRUE) &&
           requireNamespace("postie", quietly = TRUE)
knitr::opts_chunk$set(eval = can_run)
```

`blink` is a fast, deterministic **mean-field (ODE) twin** of the
[malariasimulation](https://github.com/mrc-ide/malariasimulation) individual-based
model of *Plasmodium falciparum* malaria. It reproduces the same age- and
biting-heterogeneity-structured human model and compartmental mosquito model,
takes the **same parameter list** as the IBM, and is seeded at the
[malariaEquilibrium](https://github.com/mrc-ide/malariaEquilibrium) fixed point.
A multi-decade run finishes in a couple of seconds, independent of population
size, which makes it well suited to calibration, sweeps and quick scenario work
where a stochastic IBM would be too slow.

Reach for the IBM (not this) when you need stochastic variation, individual
heterogeneity beyond the mean field, or *P. vivax* — see **Scope** at the end.

## Install and load

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("pwinskill/blink")
```

```{r load}
library(blink)
```

`malariasimulation` (for parameter lists and the `set_*` builders) and `postie`
(for post-processing) are used throughout; both are listed under Suggests.

## A basic run

Build an unmodified `malariasimulation` parameter list and run it. You supply a
target adult EIR (infectious bites per adult per year); the model seeds itself at
the corresponding equilibrium, so no separate burn-in is needed for an aseasonal
run.

```{r basic}
p <- malariasimulation::get_parameters(list(human_population = 1000))

# 15 years at a daily step, seeded at the equilibrium for EIR = 20
out <- run_simulation_ode(timesteps = 15 * 365, parameters = p, init_EIR = 20)

out[1:3, c("timestep", "n_age_730_3650", "n_detect_lm_730_3650",
           "p_detect_lm_730_3650", "EIR", "ft")]
```

The return value is a wide, `malariasimulation`-style daily count table. Column
tags carry the age band **in days**: `730_3650` is the canonical 2-10y band
(always included), and `0_36500` is all ages. `p_detect_lm_*` is LM-detectable
prevalence, `EIR` is per adult per year, and `ft` is the treated fraction. Because
the model is mean-field, prevalence and EIR are population-independent;
`human_population` only rescales the count columns.

With no interventions the model holds flat at the seeded equilibrium, so the
first and last rows agree:

```{r flat}
range(out$EIR)                    # ~20 throughout
range(out$p_detect_lm_730_3650)   # constant to machine precision
```

## Reading outputs with postie

Because the count table is `malariasimulation`-shaped, it goes straight into
[postie](https://github.com/mrc-ide/postie) — the same two calls you would make
on an IBM run, with no blink-specific wrapper in between.

```{r epi}
# prevalence: one <diagnostic>_prevalence_<lo>_<hi> column per age band (years)
prevalence <- postie::get_prevalence(out, diagnostic = "lm")
utils::tail(prevalence["lm_prevalence_2_10"], 3)

# rates: clinical / severe incidence, mortality and DALYs by age band
rates <- postie::get_rates(out)
utils::head(rates[, c("time", "age_lower", "age_upper",
                      "clinical", "severe", "dalys")])
```

Use `diagnostic = "pcr"` for PCR prevalence, and postie's own arguments
(`scaler`, `treatment_scaler`, `life_expectancy`, …) as you normally would. This
means a post-processing pipeline written for the IBM works on a `blink` run
unchanged.

### Getting finer age resolution

By default you get the 2-10y and all-ages bands. To emit more, set the
`malariasimulation` rendering fields on the parameter list before running — the
same fields the IBM uses:

```{r bands}
p_bands <- p
p_bands$prevalence_rendering_min_ages <- c(0,  2, 10) * 365
p_bands$prevalence_rendering_max_ages <- c(2, 10, 100) * 365
p_bands$clinical_incidence_rendering_min_ages <- c(0, 2) * 365
p_bands$clinical_incidence_rendering_max_ages <- c(2, 10) * 365

out_bands <- run_simulation_ode(10 * 365, p_bands, init_EIR = 20)
grep("^p_detect_lm_", names(out_bands), value = TRUE)
```

## Layering interventions

Interventions are configured with the ordinary `malariasimulation::set_*`
builders and applied automatically from the list — there are no extra arguments
to `run_simulation_ode()`. Every module starts at its baseline at t = 0, so the
equilibrium seed is preserved and each intervention acts from its scheduled
timestep. A few representative examples follow.

### Bed nets

```{r nets}
p_nets <- malariasimulation::set_bednets(
  malariasimulation::get_parameters(list(human_population = 1000)),
  timesteps = 5 * 365, coverages = 0.8, retention = 5 * 365,
  dn0 = matrix(0.387, 1, 1), rn = matrix(0.563, 1, 1),
  rnm = matrix(0.24, 1, 1), gamman = 2.64 * 365)

nets <- run_simulation_ode(12 * 365, p_nets, init_EIR = 20)
c(baseline = nets$p_detect_lm_730_3650[1],
  trough   = min(nets$p_detect_lm_730_3650))
```

Prevalence falls after deployment and then relaxes as the nets decay. Both
exponential (log-uniform) and logistic net retention are supported, and repeated
distributions accumulate correctly (net usage is the full mean-field mixture over
all past distributions).

### Seasonal SMC (needs a drug)

Chemoprevention and clinical treatment require a drug to be registered first with
`set_drugs()`. Here, four monthly SMC rounds in under-5s:

```{r smc}
p_smc <- malariasimulation::set_drugs(
  malariasimulation::get_parameters(list(human_population = 1000)),
  list(malariasimulation::SP_AQ_params))
p_smc <- malariasimulation::set_smc(
  p_smc, drug = 1, timesteps = c(365, 395, 425, 455), coverages = rep(0.9, 4),
  min_ages = rep(round(0.25 * 365), 4), max_ages = rep(round(5 * 365), 4))
p_smc$prevalence_rendering_min_ages <- round(0.25 * 365)
p_smc$prevalence_rendering_max_ages <- round(5 * 365)

smc <- run_simulation_ode(3 * 365, p_smc, init_EIR = 20)
band <- "p_detect_lm_91_1825"        # the 0.25-5y target band, tag in days
c(baseline = smc[[band]][1], trough = min(smc[[band]]))
```

SMC/MDA/PMC are applied as pulsed mass drug administration between integration
segments: a fraction (coverage x drug efficacy) of the target age band is cleared
and moved to the first stage of a chemoprevention prophylaxis chain matched to the
drug's Weibull protection curve.

### PEV vaccine (RTS,S via EPI, with a booster)

```{r pev}
p_pev <- malariasimulation::set_pev_epi(
  malariasimulation::get_parameters(list(human_population = 1000)),
  profile = malariasimulation::rtss_profile,
  timesteps = 1, coverages = 0.9, min_wait = 0, age = round(6 * 30),
  booster_spacing = round(12 * 30), booster_coverage = matrix(0.8, 1, 1),
  booster_profile = list(malariasimulation::rtss_booster_profile))

pev <- run_simulation_ode(12 * 365, p_pev, init_EIR = 20)
c(baseline = pev$p_detect_lm_730_3650[1],
  year12   = pev$p_detect_lm_730_3650[nrow(pev)])
```

PEV reduces the infection hazard by a per-age, per-time factor. Efficacy is
averaged over the per-individual antibody distribution the IBM samples (a 4-D
Gauss-Hermite rule over `cs`/`rho`/`ds`/`dl`), not evaluated at the profile
median; primary and booster doses, EPI and mass campaigns, and time-varying EPI
coverage are all modelled.
Transmission-blocking vaccines are available via `set_tbv()`.

### Custom demography

`set_demography()` supplies age-specific mortality, which reshapes the
equilibrium age structure the model is seeded at:

```{r demog}
dr <- c(0.048, 0.007, 0.003, 0.004, 0.008, 0.020, 0.050, 0.120) / 365
ag <- round(c(1, 5, 10, 20, 40, 60, 80, 100) * 365)

p_dem <- malariasimulation::set_demography(
  malariasimulation::get_parameters(list(human_population = 1000)),
  agegroups = ag, timesteps = 0, deathrates = matrix(dr, nrow = 1))
p_dem$prevalence_rendering_min_ages <- c(0, 0)
p_dem$prevalence_rendering_max_ages <- c(5, 100) * 365

dem <- run_simulation_ode(8 * 365, p_dem, init_EIR = 20)
under5_fraction <- dem$n_age_0_1825[nrow(dem)] / dem$n_age_0_36500[nrow(dem)]
under5_fraction     # ~0.09 here, vs ~0.21 under the default constant hazard
```

Custom demography is **time-varying**: `mu_age(t)` is interpolated over
`deathrate_timesteps`, so a demographic transition is modelled rather than
frozen. The baseline (`t = 0`) row additionally seeds the equilibrium age
structure. Match `default_age_lower(max_age =)` to the top `deathrate_agegroups`;
a warning fires if the model age grid runs past them.

## Seasonality and burn-in

Turn on seasonality with the Fourier rainfall parameters. Unlike an aseasonal
run, a seasonal run **oscillates** around a limit cycle rather than holding flat.
The state is seeded at the annual-mean equilibrium, so the first ~10 years are a
transient onto the cycle, and the seasonal annual-mean EIR sits a few percent
below the aseasonal `init_EIR` target (nonlinear averaging). Run long and read a
**burned-in** year for calibration or comparison:

```{r seasonal}
p_seas <- malariasimulation::get_parameters(list(
  human_population = 1000, model_seasonality = TRUE,
  g0 = 0.285, g = c(-0.33, -0.13, 0.052), h = c(-0.35, 0.020, 0.10)))

seas <- run_simulation_ode(20 * 365, p_seas, init_EIR = 20)
final_year <- seas[(nrow(seas) - 365 + 1):nrow(seas), ]
range(final_year$EIR)   # seasonal swing in the settled cycle
mean(final_year$EIR)    # a few % below the aseasonal target of 20
```

## The validation story

`blink` is checked two ways. **First, self-consistency:** with no
interventions it holds flat at the `malariaEquilibrium` seed to machine precision
(and still holds flat with treatment, `ft > 0`, thanks to a corrected
prophylaxis-aging recursion). **Second, agreement with the IBM:** the same
parameter list is passed to both models — `malariasimulation` burned in 30 years
at a population of 10,000, this model seeded at equilibrium — across six
scenarios (`comparison/`). Representative agreement:

- Equilibrium PfPR(2-10) vs EIR tracks the IBM closely (e.g. at EIR 20:
  IBM 0.553 vs ODE 0.547; at EIR 120: 0.788 vs 0.786).
- The age-prevalence profile at EIR 20 agrees to `max |IBM - ODE| = 0.008`
  (mean 0.004) across bands from infancy to 85y.
- Custom demography reproduces the IBM's under-5 population fraction
  (IBM 0.090 vs ODE 0.088).

Bed nets, seasonality and SMC scenarios are compared as well (including clinical
and severe incidence time series); see `comparison/plots/` for the figures. LM
prevalence and clinical incidence match the IBM to within ~1%. **Severe
incidence** (and DALYs derived from it) is more sensitive to the immunity model
and can differ by ~5-20% (largest at low EIR); treat it as indicative.

## Scope and approximations

- **P. falciparum only.** *P. vivax* is rejected; the model is always
  compartmental (the individual-mosquito code path does not apply).
- **Mean-field caveats.** Vector control is population-averaged (slightly
  under-suppresses at deep troughs vs the IBM); drug prophylaxis is an Erlang
  chain matched to the drug's Weibull protection (its mean and variance, not the
  Weibull itself). PCR prevalence follows the IBM convention (all D/Tr/A/U).
- **Full detail.** `vignette("model")` is the canonical specification: the ODE
  system, what each state dimension carries (and what is captured *without* a
  dimension), and an argument-by-argument table of every `malariasimulation`
  `set_*` function.

```{r session, eval = TRUE}
sessionInfo()
```
