---
title: "Getting started with blink"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with blink}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# The examples build the odin2 model and use malariasimulation + postie (both in
# Suggests). They are runnable as written; to keep package checks fast and
# dependency-light, chunks only evaluate when those packages are installed.
can_run <- requireNamespace("malariasimulation", quietly = TRUE) &&
           requireNamespace("postie", quietly = TRUE)
knitr::opts_chunk$set(eval = can_run)
```

`blink` is a fast, deterministic **mean-field (ODE) twin** of the
[malariasimulation](https://github.com/mrc-ide/malariasimulation) individual-based
model of *Plasmodium falciparum* malaria. It reproduces the same age- and
biting-heterogeneity-structured human model and compartmental mosquito model,
takes the **same parameter list** as the IBM, and is seeded at the
[malariaEquilibrium](https://github.com/mrc-ide/malariaEquilibrium) fixed point.
A multi-decade run finishes in a couple of seconds, independent of population
size, which makes it well suited to calibration, sweeps and quick scenario work
where a stochastic IBM would be too slow.

Reach for the IBM (not this) when you need stochastic variation, individual
heterogeneity beyond the mean field, or *P. vivax*; see **Scope** at the end.

## Install and load

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("pwinskill/blink")
```

```{r load}
library(blink)
```

`malariasimulation` (for parameter lists and the `set_*` builders) and `postie`
(for post-processing) are used throughout; both are listed under Suggests.

## A basic run

Build an unmodified `malariasimulation` parameter list and run it. You supply a
target adult EIR (infectious bites per adult per year); the model seeds itself at
the corresponding equilibrium, so no separate burn-in is needed for an aseasonal
run.

```{r basic}
p <- malariasimulation::get_parameters(list(human_population = 1000))

# 15 years at a daily step, seeded at the equilibrium for EIR = 20
out <- run_simulation_ode(timesteps = 15 * 365, parameters = malariasimulation::set_equilibrium(p, init_EIR = 20))

out[1:3, c("timestep", "n_age_730_3650", "n_detect_lm_730_3650",
           "p_detect_lm_730_3650", "EIR", "ft")]
```

The return value is a wide, `malariasimulation`-style daily count table. Column
tags carry the age band **in days**: `730_3650` is the canonical 2-10y band
(always included), and `0_36500` is all ages. `p_detect_lm_*` is LM-detectable
prevalence, `EIR` is per adult per year, and `ft` is the treated fraction. Because
the model is mean-field, prevalence and EIR are population-independent;
`human_population` only rescales the count columns.

With no interventions the run settles rather than holding perfectly still:

```{r flat}
range(out$EIR)                    # ~20 throughout
range(out$p_detect_lm_730_3650)   # a ~0.3% relaxation over the first years
```

Over these fifteen years LM prevalence in 2-10 year olds runs between 0.5485
(the seed, at day 0) and 0.5470 (its low point near the end of the first year),
a spread of 0.28%, and is flat to 0.02% over the last five years. EIR stays
within 0.5% of its target throughout.

That relaxation is expected, and it is not a numerical error. The seed is the
[malariaEquilibrium](https://github.com/mrc-ide/malariaEquilibrium) analytic
fixed point, and that solution encodes *simplified* forms (a linear force of
infection, refractory immunity boosting on the raw rate, exponential incubation
survival), whereas `blink` reproduces `malariasimulation`'s own per-day
semantics. The seed is therefore a close **approximation** to `blink`'s true
fixed point rather than the fixed point itself. `malariasimulation` is seeded
from the same solution and drifts off it the same way. Burn in before
calibrating, and read a settled window rather than day 0.

## Reading outputs with postie

Because the count table is `malariasimulation`-shaped, it goes straight into
[postie](https://github.com/mrc-ide/postie): the same two calls you would make
on an IBM run, with no blink-specific wrapper in between.

```{r epi}
# prevalence: one <diagnostic>_prevalence_<lo>_<hi> column per age band (years)
prevalence <- postie::get_prevalence(out, diagnostic = "lm")
utils::tail(prevalence["lm_prevalence_2_10"], 3)

# rates: clinical / severe incidence, mortality and DALYs by age band
rates <- postie::get_rates(out)
utils::head(rates[, c("time", "age_lower", "age_upper",
                      "clinical", "severe", "dalys")])
```

Use `diagnostic = "pcr"` for PCR prevalence, and postie's own arguments
(`scaler`, `treatment_scaler`, `life_expectancy`, …) as you normally would. This
means a post-processing pipeline written for the IBM works on a `blink` run
unchanged.

### Getting finer age resolution

By default you get the 2-10y and all-ages bands. To emit more, set the
`malariasimulation` rendering fields on the parameter list before running (the
same fields the IBM uses):

```{r bands}
p_bands <- p
p_bands$prevalence_rendering_min_ages <- c(0,  2, 10) * 365
p_bands$prevalence_rendering_max_ages <- c(2, 10, 100) * 365
p_bands$clinical_incidence_rendering_min_ages <- c(0, 2) * 365
p_bands$clinical_incidence_rendering_max_ages <- c(2, 10) * 365

out_bands <- run_simulation_ode(10 * 365, malariasimulation::set_equilibrium(p_bands, init_EIR = 20))
grep("^p_detect_lm_", names(out_bands), value = TRUE)
```

## Layering interventions

Interventions are configured with the ordinary `malariasimulation::set_*`
builders and applied automatically from the list. There are no extra arguments
to `run_simulation_ode()`. Every module starts at its baseline at t = 0, so the
equilibrium seed is preserved and each intervention acts from its scheduled
timestep. A few representative examples follow.

### Bed nets

```{r nets}
p_nets <- malariasimulation::set_bednets(
  malariasimulation::get_parameters(list(human_population = 1000)),
  timesteps = 5 * 365, coverages = 0.8, retention = 5 * 365,
  dn0 = matrix(0.387, 1, 1), rn = matrix(0.563, 1, 1),
  rnm = matrix(0.24, 1, 1), gamman = 2.64 * 365)

nets <- run_simulation_ode(12 * 365, malariasimulation::set_equilibrium(p_nets, init_EIR = 20))
c(baseline = nets$p_detect_lm_730_3650[1],
  trough   = min(nets$p_detect_lm_730_3650))
```

Prevalence falls after deployment and then relaxes as the nets decay. Both
exponential (log-uniform) and logistic net retention are supported, and repeated
distributions accumulate correctly (net usage is the full mean-field mixture over
all past distributions).

### Seasonal SMC (needs a drug)

Chemoprevention and clinical treatment require a drug to be registered first with
`set_drugs()`. Here, four monthly SMC rounds in under-5s:

```{r smc}
p_smc <- malariasimulation::set_drugs(
  malariasimulation::get_parameters(list(human_population = 1000)),
  list(malariasimulation::SP_AQ_params))
p_smc <- malariasimulation::set_smc(
  p_smc, drug = 1, timesteps = c(365, 395, 425, 455), coverages = rep(0.9, 4),
  min_ages = rep(round(0.25 * 365), 4), max_ages = rep(round(5 * 365), 4))
p_smc$prevalence_rendering_min_ages <- round(0.25 * 365)
p_smc$prevalence_rendering_max_ages <- round(5 * 365)

smc <- run_simulation_ode(3 * 365, malariasimulation::set_equilibrium(p_smc, init_EIR = 20))
band <- "p_detect_lm_91_1825"        # the 0.25-5y target band, tag in days
c(baseline = smc[[band]][1], trough = min(smc[[band]]))
```

SMC/MDA/PMC are applied as pulsed mass drug administration between integration
segments: a fraction (coverage x drug efficacy) of the target age band is cleared
and moved to the first stage of a chemoprevention prophylaxis chain matched to the
drug's Weibull protection curve.

### PEV vaccine (RTS,S via EPI, with a booster)

```{r pev}
p_pev <- malariasimulation::set_pev_epi(
  malariasimulation::get_parameters(list(human_population = 1000)),
  profile = malariasimulation::rtss_profile,
  timesteps = 1, coverages = 0.9, min_wait = 0, age = round(6 * 30),
  booster_spacing = round(12 * 30), booster_coverage = matrix(0.8, 1, 1),
  booster_profile = list(malariasimulation::rtss_booster_profile))

pev <- run_simulation_ode(12 * 365, malariasimulation::set_equilibrium(p_pev, init_EIR = 20))
c(baseline = pev$p_detect_lm_730_3650[1],
  year12   = pev$p_detect_lm_730_3650[nrow(pev)])
```

PEV reduces the infection hazard by a per-age, per-time factor. Efficacy is
averaged over the per-individual antibody distribution the IBM samples (a 4-D
Gauss-Hermite rule over `cs`/`rho`/`ds`/`dl`), not evaluated at the profile
median; primary and booster doses, EPI and mass campaigns, and time-varying EPI
coverage are all modelled.
Transmission-blocking vaccines are available via `set_tbv()`.

### Custom demography

`set_demography()` supplies age-specific mortality, which reshapes the
equilibrium age structure the model is seeded at:

```{r demog}
dr <- c(0.048, 0.007, 0.003, 0.004, 0.008, 0.020, 0.050, 0.120) / 365
ag <- round(c(1, 5, 10, 20, 40, 60, 80, 100) * 365)

p_dem <- malariasimulation::set_demography(
  malariasimulation::get_parameters(list(human_population = 1000)),
  agegroups = ag, timesteps = 0, deathrates = matrix(dr, nrow = 1))
p_dem$prevalence_rendering_min_ages <- c(0, 0)
p_dem$prevalence_rendering_max_ages <- c(5, 100) * 365

dem <- run_simulation_ode(8 * 365, malariasimulation::set_equilibrium(p_dem, init_EIR = 20))
under5_fraction <- dem$n_age_0_1825[nrow(dem)] / dem$n_age_0_36500[nrow(dem)]
under5_fraction     # ~0.09 here, vs ~0.21 under the default constant hazard
```

Custom demography is **time-varying**: `mu_age(t)` is interpolated over
`deathrate_timesteps`, so a demographic transition is modelled rather than
frozen. The baseline (`t = 0`) row additionally seeds the equilibrium age
structure. Match `default_age_lower(max_age =)` to the top `deathrate_agegroups`;
a warning fires if the model age grid runs past them.

## Seasonality and burn-in

Turn on seasonality with the Fourier rainfall parameters. Unlike an aseasonal
run, a seasonal run **oscillates** around a limit cycle rather than holding flat.
The state is seeded at the annual-mean equilibrium, so the first ~10 years are a
transient onto the cycle, and the seasonal annual-mean EIR sits a few percent
below the aseasonal `init_EIR` target (nonlinear averaging). Run long and read a
**burned-in** year for calibration or comparison:

```{r seasonal}
p_seas <- malariasimulation::get_parameters(list(
  human_population = 1000, model_seasonality = TRUE,
  g0 = 0.285, g = c(-0.33, -0.13, 0.052), h = c(-0.35, 0.020, 0.10)))

seas <- run_simulation_ode(20 * 365, malariasimulation::set_equilibrium(p_seas, init_EIR = 20))
final_year <- seas[(nrow(seas) - 365 + 1):nrow(seas), ]
range(final_year$EIR)   # seasonal swing in the settled cycle
mean(final_year$EIR)    # a few % below the aseasonal target of 20
```

### Solver settings

Every run above uses the default numerics. They live in `run_simulation_ode()`'s
fourth argument, `tuning`: the solver tolerances and maximum step, the age grid,
and the lengths of the Erlang lag and prophylaxis chains. Pass an `ode_tuning()`
object or a plain named list of just the fields you want to change. Nothing
epidemiological sits there (model parameters, the target EIR included, stay on
the parameter list), and every default is the validated choice, so leaving
`tuning` alone is almost always right.

The exception is a long projection like the 20-year run above, where the solver,
rather than the daily output grid, sets the step count. Loosening the relative
tolerance and letting the step grow runs about 1.4-1.7x faster on a seasonal
projection, and moves the aggregate outputs far less than anything you would
report:

```{r tuning}
seas_fast <- run_simulation_ode(
  20 * 365, malariasimulation::set_equilibrium(p_seas, init_EIR = 20),
  tuning = list(rtol = 1e-6, step_size_max = 10))

c(default = mean(final_year$EIR),
  faster  = mean(seas_fast$EIR[(nrow(seas_fast) - 365 + 1):nrow(seas_fast)]))
```

Leave `atol` at its `1e-8` default whatever else you change: the prophylaxis
chain stages hold occupancies of order `1e-6`, which a looser absolute tolerance
lets dip below zero. `?ode_tuning` documents every field.

## The validation story

`blink` is checked two ways. **First, self-consistency:** with no interventions
it stays on the `malariaEquilibrium` seed to within a few tenths of a percent and
then holds, as described under *A basic run* above (and it behaves the same way
with treatment, `ft > 0`, thanks to a corrected prophylaxis-aging recursion).
**Second, agreement with the IBM:** the same parameter list is passed to both
models, `malariasimulation` burned in 30 years at a population of 10,000 and this
model seeded at equilibrium, across eighteen scenarios (`comparison/`).
Representative agreement:

- Equilibrium PfPR(2-10) vs EIR tracks the IBM closely (at EIR 20 the IBM median
  is 0.5479 against blink's 0.5483; at EIR 120, 0.7857 against 0.7855).
- The age-prevalence profile at EIR 20 agrees to `max |IBM - blink| = 0.005`
  (mean 0.002) across twelve bands from infancy to 85y.
- Custom demography reproduces the IBM's under-5 population fraction
  (IBM 0.088 vs blink 0.088).

The other scenarios cover bed nets, IRS, seasonality, SMC, RTS,S, treatment
scale-up and fifteen-year multi-intervention programmes, with clinical and severe
incidence time series throughout; `vignette("comparison")` has the figures and
the numbers behind each of them. Across the EIR 1-120 grid LM prevalence matches
the IBM median to within 1.6% and clinical incidence to within 2.6%. **Severe
incidence** (and DALYs derived from it) is more sensitive to the immunity model
and can differ by ~5-20%, largest at EIR 20 and above and in the 5-20 year age
bands; treat it as indicative.

## Scope and approximations

- **P. falciparum only.** *P. vivax* is rejected; the model is always
  compartmental (the individual-mosquito code path does not apply).
- **Mean-field caveats.** Vector control is population-averaged (slightly
  under-suppresses at deep troughs vs the IBM); drug prophylaxis is an Erlang
  chain matched to the drug's Weibull protection (its mean and variance, not the
  Weibull itself). PCR prevalence follows the IBM convention (all D/Tr/A/U).
- **Full detail.** `vignette("model")` is the canonical specification: the ODE
  system, what each state dimension carries (and what is captured *without* a
  dimension), and an argument-by-argument table of every `malariasimulation`
  `set_*` function.

```{r session, eval = TRUE}
sessionInfo()
```
