---
title: "Getting started with fleet"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting started with fleet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# The examples build the odin2 model and use malariasimulation + postie (both in
# Suggests). They are runnable as written; to keep package checks fast and
# dependency-light, chunks only evaluate when those packages are installed.
can_run <- requireNamespace("malariasimulation", quietly = TRUE) &&
           requireNamespace("postie", quietly = TRUE)
knitr::opts_chunk$set(eval = can_run)
```

> **⚠️ Work in progress: not ready for real use.**
> `fleet` is published early so the approach, and its comparison against
> `malariasimulation`, can be examined and argued with, not so that anyone can
> rely on its numbers: the API is unstable and nothing here has been peer
> reviewed. The known discrepancies against the IBM are open rather than
> resolved: an excess on falciparum clinical and severe incidence across the
> 63-country site files that is not yet explained; a larger excess on *P. vivax*
> sites, not yet explained either; and, at school age, vivax LM prevalence and
> clinical incidence a few per cent high. The full statement
> is on the [package front page](https://pwinskill.github.io/fleet/); if you need
> results you can defend today, use
> [malariasimulation](https://github.com/mrc-ide/malariasimulation).

`fleet` is a fast, deterministic **mean-field twin** of the
[malariasimulation](https://github.com/mrc-ide/malariasimulation) individual-based
model of *Plasmodium falciparum* and *P. vivax* malaria. It reproduces the same
age- and biting-heterogeneity-structured human model and compartmental mosquito
model, advances them one day at a time in the IBM's own order, takes the **same
parameter list** as the IBM, and is seeded at the
[malariaEquilibrium](https://github.com/mrc-ide/malariaEquilibrium) solution
(for vivax, `malariaEquilibriumVivax`'s). A 30-year falciparum run takes under
two seconds and a vivax run under half a minute, independent of population
size. That speed is the point: the intended uses are calibration, sweeps and
quick scenario work, where a stochastic IBM would be too slow. It is not ready
for those jobs yet (see the note above).

Reach for the IBM (not this) when you need stochastic variation or individual
heterogeneity beyond the mean field.

## Install and load

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("pwinskill/fleet")
```

```{r load}
library(fleet)
```

`malariasimulation` (for parameter lists and the `set_*` builders) and `postie`
(for post-processing) are used throughout; both are listed under Suggests.

## A basic run

Build a `malariasimulation` parameter list and run it, exactly as you would
run the IBM. You supply a target adult EIR (infectious bites per adult per
year), and the model is seeded at the corresponding equilibrium. The list asks
for the age bands to report, as it does of the IBM: here incidence in under-5s
and at all ages, beside the default 2-10 prevalence band.

```{r basic}
p <- malariasimulation::get_parameters(list(
  human_population = 1000,
  clinical_incidence_rendering_min_ages = c(0, 0),
  clinical_incidence_rendering_max_ages = c(5, 100) * 365 - 1,
  severe_incidence_rendering_min_ages = c(0, 0),
  severe_incidence_rendering_max_ages = c(5, 100) * 365 - 1))

# 15 years at a daily step, seeded at the equilibrium for EIR = 20
out <- run_simulation_ode(timesteps = 15 * 365, parameters = malariasimulation::set_equilibrium(p, init_EIR = 20))

out[1:3, c("timestep", "n_age_730_3650", "n_detect_lm_730_3650",
           "n_inc_clinical_0_1824", "EIR")]
```

The return value is `malariasimulation`'s daily table, with the columns the IBM
would give this parameter list, under the same names and with the same
meanings. Column tags carry the age band **in days**, both ends included:
`730_3650` is the 2-10y band, and `0_1824` the under-5s. Every `n_*` column is
a count: `n_detect_lm_*` the LM-positive and `n_inc_clinical_*` the day's new
clinical cases in the band. (`p_detect_lm_*` is the IBM's expected-count
partner of `n_detect_lm_*`, not a prevalence.) `EIR` is per adult per year.
Because the model is mean-field, rates are population-independent;
`human_population` only rescales the counts.

With no interventions the run settles rather than holding perfectly still:

```{r flat}
pfpr <- out$n_detect_lm_730_3650 / out$n_age_730_3650   # LM prevalence, 2-10y
range(out$EIR)   # ~20 throughout
range(pfpr)      # a ~0.4% relaxation over the first years
```

Row 1 is day 1, whose state is the seed. Over these fifteen years LM prevalence
in 2-10 year olds runs between 0.5495 (the seed) and 0.5475 (its low point, a
little over a year in), a spread of 0.35%, and is flat to 0.008% over the last
five years. EIR stays within 0.4% of its target throughout. Incidence relaxes
further off the seed than prevalence does, a few per cent over the first years
at this EIR and more at higher ones.

That relaxation is expected, not a numerical error. The
[malariaEquilibrium](https://github.com/mrc-ide/malariaEquilibrium) seed solves
the model's continuous-time form, so it is a close **approximation** to
`fleet`'s true fixed point on the daily clock rather than the fixed point
itself; `malariasimulation` is seeded from the same solution and drifts off it
too. `vignette("model")` lists the differences. **Burn in before calibrating,
and read a settled window rather than the first days.**

## Reading outputs with postie

Because the table is `malariasimulation`'s, it goes straight into
[postie](https://github.com/mrc-ide/postie): the same two calls you would make
on an IBM run, with no fleet-specific wrapper in between.

```{r epi}
# prevalence: one <diagnostic>_prevalence_<lo>_<hi> column per age band (years)
prevalence <- postie::get_prevalence(out, diagnostic = "lm")
utils::tail(prevalence["lm_prevalence_2_10"], 3)

# rates: clinical / severe incidence, mortality and DALYs by age band
rates <- postie::get_rates(out)
utils::head(rates[, c("time", "age_lower", "age_upper",
                      "clinical", "severe", "dalys")])
```

Use `diagnostic = "pcr"` for PCR prevalence, and postie's own arguments
(`scaler`, `treatment_scaler`, `life_expectancy`, …) as you normally would. This
means a post-processing pipeline written for the IBM works on a `fleet` run
unchanged.

### Choosing age bands

The bands come from the `malariasimulation` rendering fields on the parameter
list, the same fields the IBM reads, and each family is reported over its own
list only: `prevalence_rendering_*` for detection, `clinical_incidence_*`,
`severe_incidence_*` and `incidence_rendering_*` for the three incidence
families, and `age_group_rendering_*` for population alone. A band need not
line up with the model's age groups:

```{r bands}
p_bands <- p
p_bands$prevalence_rendering_min_ages <- c(0,  2, 10) * 365
p_bands$prevalence_rendering_max_ages <- c(2, 10, 100) * 365 - 1
p_bands$clinical_incidence_rendering_min_ages <- c(0, 2) * 365
p_bands$clinical_incidence_rendering_max_ages <- c(2, 10) * 365 - 1

out_bands <- run_simulation_ode(10 * 365, malariasimulation::set_equilibrium(p_bands, init_EIR = 20))
grep("^n_detect_lm_|^n_inc_clinical_", names(out_bands), value = TRUE)
```

## Layering interventions

Interventions are configured with the ordinary `malariasimulation::set_*`
builders and applied automatically from the list. There are no extra arguments
to `run_simulation_ode()`. As in the IBM, settings read on the day (treatment
coverage, the drug mix, resistance, death rates) change on their own timestep,
and deployments (nets, spraying, vaccination, chemoprevention) act from the day
after theirs. A few representative examples follow.

### Bed nets

```{r nets}
p_nets <- malariasimulation::set_bednets(
  malariasimulation::get_parameters(list(human_population = 1000)),
  timesteps = 5 * 365, coverages = 0.8, retention = 5 * 365,
  dn0 = matrix(0.387, 1, 1), rn = matrix(0.563, 1, 1),
  rnm = matrix(0.24, 1, 1), gamman = 2.64 * 365)

nets <- run_simulation_ode(12 * 365, malariasimulation::set_equilibrium(p_nets, init_EIR = 20))
pfpr_nets <- nets$n_detect_lm_730_3650 / nets$n_age_730_3650
c(baseline = pfpr_nets[1], trough = min(pfpr_nets))
```

Prevalence falls after deployment and then relaxes as the nets decay. Both
exponential (log-uniform) and logistic net retention are supported, and repeated
distributions accumulate correctly (net usage is the full mean-field mixture over
all past distributions).

### Seasonal SMC (needs a drug)

Chemoprevention and clinical treatment require a drug to be registered first with
`set_drugs()`. Here, four monthly SMC rounds in under-5s:

```{r smc}
p_smc <- malariasimulation::set_drugs(
  malariasimulation::get_parameters(list(human_population = 1000)),
  list(malariasimulation::SP_AQ_params))
p_smc <- malariasimulation::set_smc(
  p_smc, drug = 1, timesteps = c(365, 395, 425, 455), coverages = rep(0.9, 4),
  min_ages = rep(round(0.25 * 365), 4), max_ages = rep(round(5 * 365), 4))
p_smc$prevalence_rendering_min_ages <- round(0.25 * 365)
p_smc$prevalence_rendering_max_ages <- round(5 * 365)

smc <- run_simulation_ode(3 * 365, malariasimulation::set_equilibrium(p_smc, init_EIR = 20))
pfpr_smc <- smc$n_detect_lm_91_1825 / smc$n_age_91_1825   # the 0.25-5y target band
c(baseline = pfpr_smc[1], trough = min(pfpr_smc))
```

SMC, MDA and PMC rounds are applied between days, as the IBM applies them: of
the target band, a fraction (coverage x drug efficacy) is treated. The clinical
and LM-detectable asymptomatic among them pass through the treated state first,
still detectable and infectious for a few days, and everyone treated is then
protected by a prophylaxis chain matched to the drug's Weibull protection
curve.

### PEV vaccine (RTS,S via EPI, with a booster)

```{r pev}
p_pev <- malariasimulation::set_pev_epi(
  malariasimulation::get_parameters(list(human_population = 1000)),
  profile = malariasimulation::rtss_profile,
  timesteps = 1, coverages = 0.9, min_wait = 0, age = round(6 * 30),
  booster_spacing = round(12 * 30), booster_coverage = matrix(0.8, 1, 1),
  booster_profile = list(malariasimulation::rtss_booster_profile))

pev <- run_simulation_ode(12 * 365, malariasimulation::set_equilibrium(p_pev, init_EIR = 20))
pfpr_pev <- pev$n_detect_lm_730_3650 / pev$n_age_730_3650
c(baseline = pfpr_pev[1], year12 = pfpr_pev[nrow(pev)])
```

PEV reduces the infection hazard by a per-age, per-time factor. Efficacy is
averaged over the per-individual antibody distribution the IBM samples (a 4-D
Gauss-Hermite rule over `cs`/`rho`/`ds`/`dl`), not evaluated at the profile
median; primary and booster doses, EPI and mass campaigns, and time-varying EPI
coverage are all modelled, and a campaign's protection ages with the cohort it
vaccinated.
Transmission-blocking vaccines are available via `set_tbv()`.

## Demography

`set_demography()` supplies age-specific mortality, which reshapes the
equilibrium age structure the model is seeded at:

```{r demog}
dr <- c(0.048, 0.007, 0.003, 0.004, 0.008, 0.020, 0.050, 0.120) / 365
ag <- round(c(1, 5, 10, 20, 40, 60, 80, 100) * 365)

p_dem <- malariasimulation::set_demography(
  malariasimulation::get_parameters(list(human_population = 1000)),
  agegroups = ag, timesteps = 0, deathrates = matrix(dr, nrow = 1))
p_dem$prevalence_rendering_min_ages <- c(0, 0)
p_dem$prevalence_rendering_max_ages <- c(5, 100) * 365

dem <- run_simulation_ode(8 * 365, malariasimulation::set_equilibrium(p_dem, init_EIR = 20))
under5_fraction <- dem$n_age_0_1825[nrow(dem)] / dem$n_age_0_36500[nrow(dem)]
under5_fraction     # ~0.09 here, vs ~0.21 under the default constant hazard
```

Custom demography is **time-varying**: `mu_age(t)` is interpolated over
`deathrate_timesteps`, so a demographic transition is modelled rather than
frozen. The baseline (`t = 0`) row additionally seeds the equilibrium age
structure. Match `default_age_lower(max_age =)` to the top `deathrate_agegroups`;
a warning fires if the model age grid runs past them.

## Seasonality and burn-in

Turn on seasonality with the Fourier rainfall parameters. Unlike an aseasonal
run, a seasonal run **oscillates** around a limit cycle rather than holding flat.
The state is seeded at the annual-mean equilibrium, so the first ~10 years are a
transient onto the cycle, and the seasonal annual-mean EIR sits about 7%
below the aseasonal `init_EIR` target (nonlinear averaging). Run long and read a
**burned-in** year for calibration or comparison:

```{r seasonal}
p_seas <- malariasimulation::get_parameters(list(
  human_population = 1000, model_seasonality = TRUE,
  g0 = 0.285, g = c(-0.33, -0.13, 0.052), h = c(-0.35, 0.020, 0.10)))

seas <- run_simulation_ode(20 * 365, malariasimulation::set_equilibrium(p_seas, init_EIR = 20))
final_year <- seas[(nrow(seas) - 365 + 1):nrow(seas), ]
range(final_year$EIR)   # seasonal swing in the settled cycle
mean(final_year$EIR)    # about 7% below the aseasonal target of 20
```

## P. vivax

A `get_parameters(parasite = "vivax")` list runs the *P. vivax* model:
hypnozoite batches and relapse, radical cure with liver-stage protection, and no
severe disease. The output is the IBM's vivax table, with relapses and
hypnozoite carriage beside the familiar columns:

```{r vivax}
pv <- malariasimulation::get_parameters(list(
  human_population = 1000,
  clinical_incidence_rendering_min_ages = 0,
  clinical_incidence_rendering_max_ages = 100 * 365 - 1),
  parasite = "vivax")
pv <- malariasimulation::set_drugs(pv, list(malariasimulation::CQ_PQ_params_vivax))
pv <- malariasimulation::set_clinical_treatment(pv, drug = 1, timesteps = 1, coverages = 0.4)

out_pv <- run_simulation_ode(5 * 365, malariasimulation::set_equilibrium(pv, init_EIR = 3))
tail(out_pv[, c("timestep", "n_inc_clinical_0_36499", "n_relapses",
                "n_with_hypnozoites", "iaa_mean")], 3)
```

Radical cure starts on day 1 here, from a seed without it, so these rows are
still settling towards the treated state; a level worth comparing wants a
burn-in first (`vignette("using")`). MDA, SMC and PMC are refused under vivax,
because `malariasimulation` itself fails on them.

## Discretisation settings

Every run above uses the default numerics, which live in `run_simulation_ode()`'s
fourth argument, `tuning` — an `ode_tuning()` object, or a plain named list of
just the fields you want to change. Nothing epidemiological sits there: model
parameters, the target EIR included, stay on the parameter list. There are only
three settings — the age grid, the prophylaxis chains' stage counts and the
mosquito sub-steps per day — and every default is the validated choice. The one
worth reaching for is the age grid. The default, 118 groups, is the smallest on
which every falciparum claim in `fleetcheck` passes; a grid under half as fine
runs over twice as fast, and is adequate where a result does not rest on the
level of severe incidence or on young children at high transmission:

```{r tuning}
seas_coarse <- run_simulation_ode(
  20 * 365, malariasimulation::set_equilibrium(p_seas, init_EIR = 20),
  tuning = list(age_lower = default_age_lower(n_group = 53)))

c(default = mean(final_year$EIR),
  coarse  = mean(seas_coarse$EIR[(nrow(seas_coarse) - 365 + 1):nrow(seas_coarse)]))
```

`?ode_tuning` documents every field.

## Where to go next

Four things to read before you trust a number from this model.

| | |
| --- | --- |
| **`vignette("using")`** | Where the mean field departs and what to do about it: burn-in, age grids, output bands, which settings to leave alone, which results to treat with caution, and what a run costs. |
| **[fleetcheck](https://pwinskill.github.io/fleetcheck/)** | How well it matches the IBM, as a separate project: a register of claims, each with the criterion that decides it, the value measured against it and a verdict. Not completed validation — the evidence that exists so far, with the open discrepancies named and any failing claim left failing. |
| **`vignette("model")`** | The formal specification: the daily update, and what each state dimension carries as against what is captured *without* one. |
| **`vignette("parameters")`** | Every `malariasimulation` `set_*()` function, argument by argument. |

Scope: `fleet` runs whichever parasite the list names, as `malariasimulation`
does, except for chemoprevention under vivax, which `malariasimulation` itself
cannot run. The model is always compartmental, so the individual-mosquito code
path does not apply.

```{r session, eval = TRUE}
sessionInfo()
```
