---
title: "Getting started with fleet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with fleet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# The examples build the odin2 model and use malariasimulation + postie (both in
# Suggests). They are runnable as written; to keep package checks fast and
# dependency-light, chunks only evaluate when those packages are installed.
can_run <- requireNamespace("malariasimulation", quietly = TRUE) &&
           requireNamespace("postie", quietly = TRUE)
knitr::opts_chunk$set(eval = can_run)
```

> **⚠️ Work in progress: not ready for real use.**
> `fleet` is published early so the approach, and its comparison against
> `malariasimulation`, can be examined and argued with, not so that anyone can
> rely on its numbers: the API is unstable and nothing here has been peer
> reviewed. The known discrepancies against the IBM are open rather than
> resolved, the largest being all-age severe incidence, which runs about 4 to 6%
> below the IBM, alongside a roughly 9% excess on clinical and severe incidence
> across the 63-country site files that is not yet explained. The full statement
> is on the [package front page](https://pwinskill.github.io/fleet/); if you need
> results you can defend today, use
> [malariasimulation](https://github.com/mrc-ide/malariasimulation).

`fleet` is a fast, deterministic **mean-field (ODE) twin** of the
[malariasimulation](https://github.com/mrc-ide/malariasimulation) individual-based
model of *Plasmodium falciparum* malaria. It reproduces the same age- and
biting-heterogeneity-structured human model and compartmental mosquito model,
takes the **same parameter list** as the IBM, and is seeded at the
[malariaEquilibrium](https://github.com/mrc-ide/malariaEquilibrium) fixed point.
A multi-decade run finishes in a couple of seconds, independent of population
size. That speed is the whole point of it: the intended uses are calibration,
sweeps and quick scenario work, where a stochastic IBM would be too slow. It is
not ready for those jobs yet (see the note above), so read what follows as a
description of how the package works rather than a recommendation to use it.

Reach for the IBM (not this) when you need stochastic variation, individual
heterogeneity beyond the mean field, or *P. vivax*.

## Install and load

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("pwinskill/fleet")
```

```{r load}
library(fleet)
```

`malariasimulation` (for parameter lists and the `set_*` builders) and `postie`
(for post-processing) are used throughout; both are listed under Suggests.

## A basic run

Build an unmodified `malariasimulation` parameter list and run it. You supply a
target adult EIR (infectious bites per adult per year); the model seeds itself at
the corresponding equilibrium, so no separate burn-in is needed for an aseasonal
run.

```{r basic}
p <- malariasimulation::get_parameters(list(human_population = 1000))

# 15 years at a daily step, seeded at the equilibrium for EIR = 20
out <- run_simulation_ode(timesteps = 15 * 365, parameters = malariasimulation::set_equilibrium(p, init_EIR = 20))

out[1:3, c("timestep", "n_age_730_3650", "n_detect_lm_730_3650",
           "p_detect_lm_730_3650", "EIR", "ft")]
```

The return value is a wide, `malariasimulation`-style daily count table. Column
tags carry the age band **in days**: `730_3650` is the canonical 2-10y band
(always included), and `0_36500` is all ages. `p_detect_lm_*` is LM-detectable
prevalence, `EIR` is per adult per year, and `ft` is the treated fraction. Because
the model is mean-field, prevalence and EIR are population-independent;
`human_population` only rescales the count columns.

With no interventions the run settles rather than holding perfectly still:

```{r flat}
range(out$EIR)                    # ~20 throughout
range(out$p_detect_lm_730_3650)   # a ~0.3% relaxation over the first years
```

Over these fifteen years LM prevalence in 2-10 year olds runs between 0.5485
(the seed, at day 0) and 0.5470 (its low point near the end of the first year),
a spread of 0.28%, and is flat to 0.02% over the last five years. EIR stays
within 0.5% of its target throughout.

That relaxation is expected, and it is not a numerical error. The seed is the
[malariaEquilibrium](https://github.com/mrc-ide/malariaEquilibrium) analytic
fixed point, and that solution encodes *simplified* forms (a linear force of
infection, refractory immunity boosting on the raw rate, exponential incubation
survival), whereas `fleet` reproduces `malariasimulation`'s own per-day
semantics. The seed is therefore a close **approximation** to `fleet`'s true
fixed point rather than the fixed point itself. `malariasimulation` is seeded
from the same solution and drifts off it the same way. Burn in before
calibrating, and read a settled window rather than day 0.

## Reading outputs with postie

Because the count table is `malariasimulation`-shaped, it goes straight into
[postie](https://github.com/mrc-ide/postie): the same two calls you would make
on an IBM run, with no fleet-specific wrapper in between.

```{r epi}
# prevalence: one <diagnostic>_prevalence_<lo>_<hi> column per age band (years)
prevalence <- postie::get_prevalence(out, diagnostic = "lm")
utils::tail(prevalence["lm_prevalence_2_10"], 3)

# rates: clinical / severe incidence, mortality and DALYs by age band
rates <- postie::get_rates(out)
utils::head(rates[, c("time", "age_lower", "age_upper",
                      "clinical", "severe", "dalys")])
```

Use `diagnostic = "pcr"` for PCR prevalence, and postie's own arguments
(`scaler`, `treatment_scaler`, `life_expectancy`, …) as you normally would. This
means a post-processing pipeline written for the IBM works on a `fleet` run
unchanged.

### Getting finer age resolution

By default you get the 2-10y and all-ages bands. To emit more, set the
`malariasimulation` rendering fields on the parameter list before running (the
same fields the IBM uses):

```{r bands}
p_bands <- p
p_bands$prevalence_rendering_min_ages <- c(0,  2, 10) * 365
p_bands$prevalence_rendering_max_ages <- c(2, 10, 100) * 365
p_bands$clinical_incidence_rendering_min_ages <- c(0, 2) * 365
p_bands$clinical_incidence_rendering_max_ages <- c(2, 10) * 365

out_bands <- run_simulation_ode(10 * 365, malariasimulation::set_equilibrium(p_bands, init_EIR = 20))
grep("^p_detect_lm_", names(out_bands), value = TRUE)
```

## Layering interventions

Interventions are configured with the ordinary `malariasimulation::set_*`
builders and applied automatically from the list. There are no extra arguments
to `run_simulation_ode()`. Every module starts at its baseline at t = 0, so the
equilibrium seed is preserved and each intervention acts from its scheduled
timestep. A few representative examples follow.

### Bed nets

```{r nets}
p_nets <- malariasimulation::set_bednets(
  malariasimulation::get_parameters(list(human_population = 1000)),
  timesteps = 5 * 365, coverages = 0.8, retention = 5 * 365,
  dn0 = matrix(0.387, 1, 1), rn = matrix(0.563, 1, 1),
  rnm = matrix(0.24, 1, 1), gamman = 2.64 * 365)

nets <- run_simulation_ode(12 * 365, malariasimulation::set_equilibrium(p_nets, init_EIR = 20))
c(baseline = nets$p_detect_lm_730_3650[1],
  trough   = min(nets$p_detect_lm_730_3650))
```

Prevalence falls after deployment and then relaxes as the nets decay. Both
exponential (log-uniform) and logistic net retention are supported, and repeated
distributions accumulate correctly (net usage is the full mean-field mixture over
all past distributions).

### Seasonal SMC (needs a drug)

Chemoprevention and clinical treatment require a drug to be registered first with
`set_drugs()`. Here, four monthly SMC rounds in under-5s:

```{r smc}
p_smc <- malariasimulation::set_drugs(
  malariasimulation::get_parameters(list(human_population = 1000)),
  list(malariasimulation::SP_AQ_params))
p_smc <- malariasimulation::set_smc(
  p_smc, drug = 1, timesteps = c(365, 395, 425, 455), coverages = rep(0.9, 4),
  min_ages = rep(round(0.25 * 365), 4), max_ages = rep(round(5 * 365), 4))
p_smc$prevalence_rendering_min_ages <- round(0.25 * 365)
p_smc$prevalence_rendering_max_ages <- round(5 * 365)

smc <- run_simulation_ode(3 * 365, malariasimulation::set_equilibrium(p_smc, init_EIR = 20))
band <- "p_detect_lm_91_1825"        # the 0.25-5y target band, tag in days
c(baseline = smc[[band]][1], trough = min(smc[[band]]))
```

SMC/MDA/PMC are applied as pulsed mass drug administration between integration
segments: a fraction (coverage x drug efficacy) of the target age band is cleared
and moved to the first stage of a chemoprevention prophylaxis chain matched to the
drug's Weibull protection curve.

### PEV vaccine (RTS,S via EPI, with a booster)

```{r pev}
p_pev <- malariasimulation::set_pev_epi(
  malariasimulation::get_parameters(list(human_population = 1000)),
  profile = malariasimulation::rtss_profile,
  timesteps = 1, coverages = 0.9, min_wait = 0, age = round(6 * 30),
  booster_spacing = round(12 * 30), booster_coverage = matrix(0.8, 1, 1),
  booster_profile = list(malariasimulation::rtss_booster_profile))

pev <- run_simulation_ode(12 * 365, malariasimulation::set_equilibrium(p_pev, init_EIR = 20))
c(baseline = pev$p_detect_lm_730_3650[1],
  year12   = pev$p_detect_lm_730_3650[nrow(pev)])
```

PEV reduces the infection hazard by a per-age, per-time factor. Efficacy is
averaged over the per-individual antibody distribution the IBM samples (a 4-D
Gauss-Hermite rule over `cs`/`rho`/`ds`/`dl`), not evaluated at the profile
median; primary and booster doses, EPI and mass campaigns, and time-varying EPI
coverage are all modelled.
Transmission-blocking vaccines are available via `set_tbv()`.

### Custom demography

`set_demography()` supplies age-specific mortality, which reshapes the
equilibrium age structure the model is seeded at:

```{r demog}
dr <- c(0.048, 0.007, 0.003, 0.004, 0.008, 0.020, 0.050, 0.120) / 365
ag <- round(c(1, 5, 10, 20, 40, 60, 80, 100) * 365)

p_dem <- malariasimulation::set_demography(
  malariasimulation::get_parameters(list(human_population = 1000)),
  agegroups = ag, timesteps = 0, deathrates = matrix(dr, nrow = 1))
p_dem$prevalence_rendering_min_ages <- c(0, 0)
p_dem$prevalence_rendering_max_ages <- c(5, 100) * 365

dem <- run_simulation_ode(8 * 365, malariasimulation::set_equilibrium(p_dem, init_EIR = 20))
under5_fraction <- dem$n_age_0_1825[nrow(dem)] / dem$n_age_0_36500[nrow(dem)]
under5_fraction     # ~0.09 here, vs ~0.21 under the default constant hazard
```

Custom demography is **time-varying**: `mu_age(t)` is interpolated over
`deathrate_timesteps`, so a demographic transition is modelled rather than
frozen. The baseline (`t = 0`) row additionally seeds the equilibrium age
structure. Match `default_age_lower(max_age =)` to the top `deathrate_agegroups`;
a warning fires if the model age grid runs past them.

## Seasonality and burn-in

Turn on seasonality with the Fourier rainfall parameters. Unlike an aseasonal
run, a seasonal run **oscillates** around a limit cycle rather than holding flat.
The state is seeded at the annual-mean equilibrium, so the first ~10 years are a
transient onto the cycle, and the seasonal annual-mean EIR sits a few percent
below the aseasonal `init_EIR` target (nonlinear averaging). Run long and read a
**burned-in** year for calibration or comparison:

```{r seasonal}
p_seas <- malariasimulation::get_parameters(list(
  human_population = 1000, model_seasonality = TRUE,
  g0 = 0.285, g = c(-0.33, -0.13, 0.052), h = c(-0.35, 0.020, 0.10)))

seas <- run_simulation_ode(20 * 365, malariasimulation::set_equilibrium(p_seas, init_EIR = 20))
final_year <- seas[(nrow(seas) - 365 + 1):nrow(seas), ]
range(final_year$EIR)   # seasonal swing in the settled cycle
mean(final_year$EIR)    # a few % below the aseasonal target of 20
```

### Solver settings

Every run above uses the default numerics. They live in `run_simulation_ode()`'s
fourth argument, `tuning`: the solver tolerances and maximum step, the age grid,
and the lengths of the Erlang lag and prophylaxis chains. Pass an `ode_tuning()`
object or a plain named list of just the fields you want to change. Nothing
epidemiological sits there (model parameters, the target EIR included, stay on
the parameter list), and every default is the validated choice, so leaving
`tuning` alone is almost always right.

The exception is a long projection like the 20-year run above, where the solver,
rather than the daily output grid, sets the step count. Loosening the relative
tolerance runs about 1.4-1.7x faster on a seasonal projection, and moves the
aggregate outputs far less than anything you would report. It is `rtol` that buys
this: `step_size_max` is a safety rail against a trial step overshooting an
interpolation grid, and raising it changes neither the step count nor the
outputs.

```{r tuning}
seas_fast <- run_simulation_ode(
  20 * 365, malariasimulation::set_equilibrium(p_seas, init_EIR = 20),
  tuning = list(rtol = 1e-6, step_size_max = 10))

c(default = mean(final_year$EIR),
  faster  = mean(seas_fast$EIR[(nrow(seas_fast) - 365 + 1):nrow(seas_fast)]))
```

Leave `atol` at its `1e-8` default whatever else you change: the prophylaxis
chain stages hold occupancies of order `1e-6`, which a looser absolute tolerance
lets dip below zero. `?ode_tuning` documents every field.

## Where to go next

This has been a tour of the interface. Three things are worth reading before you
trust a number that comes out of it.

**How well it matches the IBM.** Every claim about agreement, with the figures
and the numbers behind them, is in `vignette("comparison")`: the same parameter
list through both models across eighteen scenarios, plus a 63-country site-file
comparison. None of it is completed validation; it is the evidence that exists so
far, and the open discrepancies are named rather than buried.

**Where the mean field departs, and what to do about it.**
`vignette("using")` is the practical companion to this one: burn-in, age grids,
output bands, which settings to leave alone, which results to treat with caution,
and what a run costs.

**The formal specification.** `vignette("model")` has the ODE system, what each
state dimension carries (and what is captured *without* one), and an
argument-by-argument table of every `malariasimulation` `set_*()` function.

One scope note that belongs here rather than there: `fleet` is **P. falciparum
only**. *P. vivax* parameter lists are rejected at input, and the model is always
compartmental, so the individual-mosquito code path does not apply.

```{r session, eval = TRUE}
sessionInfo()
```
